HO unification from object language to meta language

### 2024-03-23 paper/main.tex
url
w

\paragraph{Contributions}
In this paper we discuss alternative encodings of Coq in
Elpi (Section~\ref{sec:encodings}), then we identify a minimal language \Fo{}
in which the problems sketched here can be fully described.
We then detail an encoding \elpiIn{comp} from \Fo{} to \Ho (the language of
the meta language) and a decoding \elpiIn{decomp} to relate the unifiers
bla bla.. TODO citare Teyjus.

@@

\subsection{The intuition in a nutshell}
\label{sec:nutshell}
A term $s$ is compiled in a term $t$ where any
``problematic'' subterm $p$ is replaced by a fresh unification variable $h$
and an accessory link that represent a suspended unification problem
$h \Ue p$. As a result \Ue is ``well behaved'' on $t$, meaning it does not
contradict \Eo (as it would do on ``problematic'' terms).
We now define ``problematic'' and ``well behaved'' more formally.

@@

\noindent
An example of term $t$ in \maybeeta{} is
$\lambda x.\lambda y.F~y~x$
since the substitution
$\rho = \{ F \mapsto \lambda a.\lambda b.fba\}$
makes $\rho t = \lambda x.\lambda y.f x y$
that is the eta long form of $f$.

@@

\newcommand{\subterm}[1]{\ensuremath{\mathcal{P}(#1)}}
\begin{definition}[Subterm \subterm{t}]
$$
\begin{array}{l}
t \in \subterm{t}\\
t = f t_1\ldots t_n \Rightarrow \subterm{t_i} \subseteq \subterm{t} \land f \in \subterm{t}\\
t = \lambda x.t' \Rightarrow \subterm{t'} \subseteq \subterm{t}\\
\end{array}
$$
\end{definition}

\noindent
We write $\subterm{X} = \bigcup_{t\in X} \subterm{t}$.

@@

Note that proposition \ref{prop:nf} does not hold for \Uo{} since
decompilation can introduce (actually restore) terms in
\maybeeta or \maybebeta that were move out of the way (put in $\linkStore$)
by compilation.

@@

\noindent
In the case of \Fo{} unification variables \elpiIn{fuva} have no explicit scope:
the arguments of an higher order unification variable are via the \elpiIn{fapp}
constructor. For example in the statement of the instance \coqIn{forall_dec} the
term \coqIn{P x} is represented as \elpiIn{fapp[fuva N, x]}, where \elpiIn{N} is
a memory address and \elpiIn{x} is a bound variable.

In \Ho the representation of \coqIn{P x} is instead \elpiIn{uva N [x]}.
We say that the unification variable \elpiIn{uva N L} is in \llambda iff
\elpiIn{distinct L} holds.

\begin{elpicode}
type distinct list A -> o.
distinct [].
distinct [X|XS] :- name X, not(mem X XS), distinct XS.
\end{elpicode}

\noindent
The \elpiIn{name} builtin predicate tests if a term is a bound variable.
\footnote{one could always load name x for every x under a pi and get rid of the name builtin}
The
compiler~\ref{sec:xxx} needs to support terms outside \llambda for practical
reasons, so we don't assume all out terms are in \llambda but rather test. \textbf{\textcolor{red}{what?? }}

In both languages unification variables are identified by a natural number,
which can be seen as a memory address. The memory and its associated
operations are described below:

@@

\noindent 
If a memory cell is \elpiIn{none}, then the corresponding unification variable
is not set. \elpiIn{assign} sets an unset cell to the given value.

Since in \Ho unification variables have a scope, their solution needs to be
abstracted over it in order to enable the instantiation of a single
solution to different scopes. On the contrary solutions to \Fo variables
are plain terms.

@@

\begin{invariant}[Unification variable arity] Each variable \elpiIn{A}
in \Ho has a (unique) arity \elpiIn{N} and each occurrence~
\elpiIn{(uvar A L)} is such that \elpiIn{(len L N)} holds
\end{invariant}

>>>>

\paragraph{Contributions}
In this paper we discuss alternative encodings of Coq in
Elpi (Section~\ref{sec:encodings}), then we identify a minimal language \Fo{}
in which the problems sketched here can be fully described.
We then detail an encoding \elpiIn{comp} from \Fo{} to \Ho (the language of
the meta language) and a decoding \elpiIn{decomp} to relate the unifiers
bla bla.. TODO citare Teyjus.
The code discussed in the paper can be accessed at the URL:
\url{https://github.com/FissoreD/paper-ho}.

@@

\subsection{The intuition in a nutshell}
\label{sec:nutshell}
A term $s$ is compiled in a term $t$ where every
``problematic'' sub term $p$ is replaced by a fresh unification variable $h$
and an accessory link that represent a suspended unification problem
$h \Ue p$. As a result \Ue is ``well behaved'' on $t$, that is it does not
contradict \Eo as it would otherwise do on ``problematic'' terms.
We now define ``problematic'' and ``well behaved'' more formally.

@@

\noindent
An example of term $t$ in \maybeeta{} is
$\lambda x.\lambda y.F~y~x$
since the substitution
$\rho = \{ F \mapsto \lambda a.\lambda b.fba\}$
makes $\rho t = \lambda x.\lambda y.f x y$
that is the eta long form of $f$. This term is problematic since
its rigid part, the $\lambda$-abstractions, cannot justify a
unification failure against, say, a constant.

@@

\newcommand{\subterm}[1]{\ensuremath{\mathcal{P}(#1)}}
\begin{definition}[Subterms \subterm{t}] The set of sub terms of $t$ is the
  largest set $\mathcal{P(t)}$ that can be obtained by the following rules.
$$
\begin{array}{l}
t \in \subterm{t}\\
t = f t_1\ldots t_n \Rightarrow \subterm{t_i} \subseteq \subterm{t} \land f \in \subterm{t}\\
t = \lambda x.t' \Rightarrow \subterm{t'} \subseteq \subterm{t}\\
\end{array}
$$
\end{definition}

\noindent
We write $\subterm{X} = \bigcup_{t\in X} \subterm{t}$ when $X$ is a set of terms.

@@

Note that proposition \ref{prop:nf} does not hold for \Uo{} since
decompilation can introduce (actually restore) terms in
\maybeeta or \maybebeta that were move out of the way (put in $\linkStore$)
during compilation.

@@

\noindent
Unification variables (\elpiIn{fuva} term constructor)
in \Fo{} have no explicit scope:
the arguments of an higher order variable are given via the \elpiIn{fapp}
constructor. For example the term \coqIn{P x} is represented as
\elpiIn{fapp[fuva N, x]}, where \elpiIn{N} is a memory address and
\elpiIn{x} is a bound variable.\\
In \Ho the representation of \coqIn{P x} is instead \elpiIn{uva N [x]},
since unification variables come equipped with an explicit scope.
We say that the unification variable occurrence \elpiIn{uva N L} is in
\llambda if and only if \elpiIn{L} is made of distinct names. The
predicate to test this condition is called \elpiIn{pattern-fragment}:

\input{code/pattern_fragment}

\noindent
The \elpiIn{name} builtin predicate tests if a term is a bound variable.
\footnote{one could always load name x for every x under a pi and get rid of the name builtin}

In both languages unification variables are identified by a natural number
representing a memory address.
The memory and its associated operations are described below:

@@

\noindent 
If a memory cell is \elpiIn{none}, then the corresponding unification variable
is not set. \elpiIn{assign} sets an unset cell to the given value, while
\elpiIn{new} finds the first unused address and sets it to \elpiIn{none}.\marginpar{is new used?}

Since in \Ho unification variables have a scope, their solution needs to be
abstracted over it to enable the instantiation of a single
solution to different scopes.
On the contrary a solution to a \Fo variable is a plain term.

@@

\begin{invariant}[Unification variable arity] Each variable \elpiIn{A}
in \Ho has a (unique) arity \elpiIn{N} and each occurrence~
\elpiIn{(uva A L)} is such that \elpiIn{(len L N)} holds
\end{invariant}
